Illusion Sensitivity and Phenomenological Control

### 2025-05-20 analysis/1_cleaning.qmd
simplify outlier handling in IG
@AnafNeves

```{r}
#| echo: true
#| code-fold: false

outliers <- list()
outliers_half <- list()
```

## Reaction Time (per Block)

**Note**. If the RT is collapsed to the left (i.e., has > 1/3 of ultra-fast responses - typically < 200 ms) in the first set, the entire participant will be discarded (suggesting that they did not properly do the task), but if only the second set is bad, then only the second set will be discarded (as the illusion sensitivity can still be estimated, albeit with less precision).

@@

errorrate <- data_ig |>
  group_by(Participant, Illusion_Type, Block) |>
  summarize(ErrorRate_per_block = sum(Error) / n()) |>
  ungroup() |>
  arrange(desc(ErrorRate_per_block))

d_all <- estimate_density(filter(data_ig, RT < 3)$RT)

plot_distribution <- function(dat) {

  data_error <- errorrate |>
    filter(Participant %in% unique(dat$Participant)) |>
    group_by(Participant, Block) |>
    summarize(y = mean(ErrorRate_per_block), .groups="drop") |>
    mutate(x = ifelse(Block == "A", 2.1, 2.3),
           color = case_when(
              Participant %in% outliers ~ "red",
              Participant %in% outliers_half ~ "orange",
              TRUE ~ "blue"
            ))

  dat |>
    filter(RT < 3) |>
    estimate_density(select = "RT", at = c("Participant", "Block")) |>
    group_by(Participant) |>
    normalize(select = "y") |>
    ungroup() |>
    mutate(
      Participant = fct_relevel(Participant, sort(unique(dat$Participant))),
      color = case_when(
        Participant %in% outliers ~ "red",
        Participant %in% outliers_half ~ "orange",
        TRUE ~ "blue"
      )
    ) |>
    ggplot(aes(x = x, y = y)) +
    geom_bar(data = data_error, aes(fill = color), stat = "identity", width=0.19) +
    geom_segment(aes(x = 2, xend = 2.4, y = 0.5, yend = 0.5), color = "black", linetype="dashed", linewidth = 0.5) +
    geom_area(data = normalize(d_all, select = "y"), alpha = 0.2) +
    geom_line(aes(color = color, group = interaction(Participant, Block), linetype = Block), linewidth = 0.8) +
    # geom_vline(xintercept = 0.125, linetype = "dashed", color = "red", size = 0.5) +
    scale_color_manual(values = c("red" = "#F44336", "orange" = "#FF9800", "blue" = "blue"), guide = "none") +
    scale_fill_manual(values = c("red" = "#F44336", "orange" = "#FF9800", "blue" = "blue"), guide = "none") +
    scale_x_continuous(expand = c(0, 0), breaks=c(0, 0.5, 1, 1.5, 2), labels=c("0", "0.5", "1", "1.5", "2")) +
    scale_y_continuous(expand = c(0, 0)) +
    coord_cartesian(xlim = c(0, 2.4)) +
    theme_modern() +
    theme(axis.text.y = element_blank(),
          axis.line.y = element_blank(),
          strip.text.x = element_text(size = rel(1.5)),
          legend.position = "none") +
    facet_wrap(~Participant, nrow=10) +
    labs(y = "", x = "Reaction Time (s)")
}
```

### 1-100

```{r}
#| fig-width: 15
#| fig-height: 15
#| eval: false

plot_distribution(data_ig[data_ig$Participant %in% data_ppt$Participant[1:100],])
```

### 101-200

```{r}
#| fig-width: 15
#| fig-height: 15
#| eval: false

plot_distribution(data_ig[data_ig$Participant %in% data_ppt$Participant[1:200],])

:::

We discarded `r length(outliers)` participants (entirely) and `r length(outliers_half)` participant's second blocks.

@@

## Error Rate (per Block)

For each block, we computed the error rate and, if more than 50%, we discarded the whole block (as it likely indicates that instructions got mixed up, for instance participants were selecting the smaller instead of the bigger circle).

```{r p_outliers_errorblock}
#| eval: false

errorrate |>
  estimate_density(at = c("Illusion_Type", "Block"), method = "KernSmooth") |>
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(color = Illusion_Type, linetype = Block)) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  scale_x_continuous(expand = c(0, 0), labels = scales::percent) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_color_manual(values = c("Ebbinghaus" = "#2196F3", "MullerLyer" = "#4CAF50", "VerticalHorizontal" = "#FF5722")) +
  labs(y = "Distribution", x = "Error Rate") +
  theme_modern()

remove_badblocks <- function(df) {
  n <- nrow(df)
  df <- df |>
    group_by(Participant, Illusion_Type, Block) |>
    mutate(ErrorRate_per_block = sum(Error) / n()) |>
    ungroup() |>
    filter(ErrorRate_per_block < 0.5) |>
    select(-ErrorRate_per_block)

  text <- paste0(
    "We removed ",
    n - nrow(df),
    " (",
    insight::format_value((n - nrow(df)) / n, as_percent = TRUE),
    ") trials belonging to bad blocks."
  )
  list(data = df, text = text)
}

out <- remove_badblocks(data_ig)
print(paste("Illusion (session 1):", out$text))
data_ig <- out$data
```

## Reaction Time (per Block)

**Note**. The removal of individual trials will also be performed [RT < 200 ms or > 3 SD; following Thériault et al. (2024)]. To mitigate the risk of confounding effects driven by extreme speed or accuracy strategies, participants whose RTs are significantly slower than the group average (RT > 4 SD above the mean, based on Makowski et al., 2023) will be excluded from the analysis.

```{r outliers_RTtrials}
#| eval: false

check_trials <- function(df) {
  data <- df |>
    mutate(Outlier = ifelse(RT >= 10, TRUE, FALSE)) |>
    group_by(Participant) |>
    mutate(Outlier = ifelse(RT < 0.150 | standardize(RT, robust = TRUE) > 4, TRUE, Outlier)) |>
    ungroup()

  p1 <- data |>
    filter(RT < 10) |>
    estimate_density(select = "RT", at = "Participant") |>
    group_by(Participant) |>
    normalize(select = "y") |>
    ungroup() |>
    merge(data |>
      group_by(Participant) |>
      mutate(Threshold = median(RT) + 4 * mad(RT)) |>
      filter(Error == 0) |>
      summarize(Threshold = mean(Threshold))) |>
    mutate(Outlier = ifelse(x >= Threshold, TRUE, FALSE)) |>
    ggplot(aes(x = x, y = y)) +
    geom_area(data = normalize(estimate_density(filter(data, RT < 10), select = "RT"), select = "y"), alpha = 0.2) +
    geom_line(aes(color = Participant, linetype = Outlier), alpha = 0.2) +
    geom_vline(xintercept = c(125), linetype = "dashed", color = "red") +
    scale_color_material_d("rainbow", guide = "none") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    guides(linetype = "none") +
    coord_cartesian(xlim = c(0, 5)) +
    theme_modern() +
    theme(axis.text.y = element_blank()) +
    labs(y = "", x = "Reaction Time (s)")

  p2 <- data |>
    group_by(Participant) |>
    summarize(Outlier = sum(Outlier) / n()) |>
    mutate(Participant = fct_reorder(Participant, Outlier)) |>
    ggplot(aes(x = Participant, y = Outlier)) +
    geom_bar(stat = "identity", aes(fill = Participant)) +
    scale_fill_material_d("rainbow", guide = "none") +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
    see::theme_modern() +
    theme(axis.text.x = element_blank()) +
    labs(y = "Percentage of outlier trials")

  text <- paste0(
    "We removed ",
    sum(data$Outlier),
    " (",
    insight::format_value(sum(data$Outlier) / nrow(data), as_percent = TRUE),
    ") outlier trials (150 ms < RT < 4 MAD above median)."
  )

  data <- filter(data, Outlier == FALSE)
  data$Outlier <- NULL

  list(p = p1 / p2, data = data, text = text)
}
```

```{r p_outliers_RTtrials}
#| results: 'asis'
#| eval: false

# out <- check_trials(data_ig)
# out$text
# out$p
# data_ig <- out$data
```

## Manipulation/Attention Checks

>>>>

## Illusion Game Participants

We removed participants with an error rate > 45% (suggesting chance-level, i.e., random, responding) for any of the illusion types. Outliers detected in the first block will be discarded entirely (suggesting that they did not properly do the task), but if only the second set is bad, then only the second set will be discarded (suggesting that they got "bored" when having to re-do the task), as the illusion sensitivity can still be estimated, albeit with less precision. Additionally, participants whose RTs are significantly slower than the group average (possibly indicative of a different strategy) were also excluded from the analysis (RT > 4 SD above the mean).

@@

data_outliers <- data_ig |>
  summarize(p_Error = sum(Error) / n(),
            z_RT = mean(RT),
            .by = c("Participant", "Illusion_Type", "Block")) |> 
  mutate(z_RT = as.numeric(standardize(z_RT)), .by = c("Illusion_Type", "Block")) |>
  arrange(desc(p_Error))

outliers <- filter(data_outliers, p_Error > 0.45 | z_RT > 4)
outliers_full <- unique(filter(outliers, Block == "A")$Participant)
outliers_half <- unique(filter(outliers, Block == "B")$Participant)
outliers_half <- outliers_half[!(outliers_half %in% outliers_full)]

p1 <- data_outliers |> 
  ggplot(aes(x = p_Error)) +
  geom_density(aes(color=Illusion_Type)) +
  geom_vline(xintercept = 0.45, linetype = "dashed", color="red")

p2 <- data_outliers |> 
  ggplot(aes(x = z_RT)) +
  geom_density(aes(color=Illusion_Type)) +
  geom_vline(xintercept = 4, linetype = "dashed", color="red")

p1 / p2
```

We discarded `r length(outliers_full)` participants (entirely) and `r length(outliers_half)` participant's second blocks based on error rate.

@@

```{r}
data_outliers <- df |>
  mutate(Outlier = ifelse(RT >= 10, TRUE, FALSE)) |>
  group_by(Participant) |>
  mutate(Outlier = ifelse(RT < 0.150 | standardize(RT, robust = TRUE) > 4, TRUE, Outlier)) |>
  ungroup()
```

## Illusion Game Trials

The removal of individual trials will also be performed (RT < 0.2 s or > 5 s; Thériault et al., 2024) to mitigate the risk of confounding effects driven by extreme speed or accuracy strategies.

```{r}
data_ig <- filter(data_ig, RT > 0.2 & RT < 5)
```

## PCS
